Dialect: A Compact Line Notation Based on SMILES

### 2022-06-09 manuscript.tex
Replaces molecular graph with molecular tree.

\documentclass{article}
\usepackage{caption}
\usepackage{algorithm2e}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{array}
\usepackage{courier}
\usepackage{csquotes}
\usepackage[USenglish]{babel}
\usepackage{nameref}

@@

Dialect's purpose is to solve the \textit{molecular serialization} problem. Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction. This is useful because it enables general-purpose devices and networks to convey complex pieces of domain-specific information. Receiver and recipient may be separated by time, space, or both. Molecular serialization extends the concept of serialization to molecular structure. Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile ("molfile")~\cite{ctfileFormats}.

Molecular serialization require careful attention to two complementary rule sets: \textit{syntax} and \textit{semantics}. Syntax defines the set of valid strings for a language, whereas semantics determine the meaning of valid strings. Incomplete specification of either syntax or semantics can lead to data loss and corruption. The semantics for Dialect are complex due to the brevity of its syntax and the inherently complex nature of molecular representation.

As a molecular serialization scheme, Dialect must balance two opposing factors: generality and complexity. If the chemical space Dialect spans is too broad, complete implementation could become impractical due to language complexity. But if the language is made too simple, the chemical space it spans may be of little interest. An important goal for Dialect is therefore to address the most interesting chemical space with the simplest possible syntax and semantics.

Toward this end, Dialect is built on four mutually-exclusive concepts:

\begin{itemize}
\item \textit{Constitution}, the atoms present in a molecule and how they are connected with each other through bonds. Dialect supports elements having two-letter IUPAC designations and all bonding arrangements available through the valence bond model.
\item \textit{Conformation}, the kind of stereoisomerism resulting from restricted rotation about a bond. Dialect support restricted rotation about double bonds leading to alkene isomerism.
\item \textit{Configuration}, the kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space. Dialect supports configurations for atoms having tetrahedral symmetry attached to four neighboring atoms.
\item \textit{Delocalization}, a type of bonding in which electrons are shared among three or more atoms, or across a combination of atoms and bonds. Dialect supports a limited form of delocalization operating over paths of alternating single and double bonds.
\end{itemize}

@@

Dialect molecules inherit behavior from an underlying \textit{molecular tree}. A tree is a data structure comprised of nodes. Each node is associated with zero or more child nodes through a directed edge. A molecular tree is a specialized tree onto which chemically-relevant metadata have been overlaid. The presence of this overlay allows nodes and edges to represent atoms and bonds, respectively.

A molecular tree is related to the more widely-used concept of a \textit{molecular graph}~\cite{balaban:1985}. There are two main difference: (1) bonds are undirected in molecular graphs but directed in molecular trees; and (2) cycles are possible in molecular graphs but not in molecular trees. The difference might appear to render trees unsuitable to represent molecules, in which cycles often play a prominent structural role. But as the next sections demonstrate, the differences between molecular trees and graphs are not so large as they might at first appear.

A Dialect molecule is a molecular tree having zero or more nodes and zero or more edges. The empty molecule, devoid of atoms and bonds, is therefore allowed. Dialect imposes no upper bound on the number of nodes or edges, although limitations in hardware and software may impose a practical upper bound on molecular size.

\section*{Data Model}

Dialect uses several abstractions collectively known as the \textit{data model}. The data model defines metadata attributes associated with the components of a molecular tree. These attributes are themselves composed of \textit{primitives} (Table~\ref{table:primitives}), or irreducible data types.

% TODO: completely re-work this to fit with the exact model being used by the reference implementation

\begin{table}
\begin{minipage}{\textwidth}
\renewcommand*\footnoterule{}
\caption{Data Model Primitives.}
\label{table:primitives}
\centering
\begin{tabular}{l l l}
    \hline
    Name & Description & Values\footnote{Ranges defined as \ttt{a..b} include \ttt{a} and exclude \ttt{b}.} \\
    \hline
    \ttt{Option<T>} & Optional Value & \ttt{None}, \ttt{T} \\
    \ttt{Three} & Integer count & 1..4 \\
    \ttt{Dec} & Integer count &0..10 \\
    \ttt{SignedDec} & Integer count & -9..10 \\
    \ttt{Thou} & Integer count & 1..1000 \\
    \ttt{Element} & IUPAC-\-approved element symbol & \ttt{Ac}, \ttt{Ag}, \ttt{Al}, {\ldots}\ttt{Zn}\\
    \ttt{Configuration} & Configurational descriptor & \ttt{Right}, \ttt{Left} \\
    \ttt{Direction} & Conformational descriptor & \ttt{Up}, \ttt{Down} \\
    \ttt{Implicit} & Implicit hydrogen count flag & - \\
    \ttt{boolean} & Boolean & \ttt{true}, \ttt{false} \\
    \hline
\end{tabular}
\end{minipage}
\end{table}

An atom possesses several attributes (Table~\ref{table:atomic-state}), which collectively define an \textit{atomic state}. Not all atomic states are valid. Implementations must ensure either the impossibility of an invalid state, or an error condition in the event that one is created. Atomic state is restricted in the following ways:

\begin{table}
\caption{Atomic State.}
\centering
\begin{tabular}{l L L l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{element} & Elemental symbol & \ttt{Option} \ttt{<Element>} & \ttt{None} \\
    \ttt{hydrogens} & Hydrogen count & \ttt{Implicit}, \ttt{Dec} & 0 \\
    \ttt{configuration} & Configuration & \ttt{Option} \ttt{<Configuration>} & \ttt{None} \\
    \ttt{charge} & Formal charge & \ttt{SignedDec} & 0 \\
    \ttt{isotope} & Mass number & \ttt{Option<Thou>} & \ttt{None} \\
    \ttt{selected} & Selection status & \ttt{boolean} & \ttt{false} \\
    \hline
\end{tabular}
\label{table:atomic-state}
\end{table}

\begin{enumerate}
    \item{If the value of \ttt{hydrogens} equals \ttt{Implicit}, then the values of \ttt{isotope}, \ttt{configuration}, and \ttt{charge} must equal their respective defaults. Furthermore, the value of \ttt{element} must equal one of \ttt{B}; \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; \ttt{S}; \ttt{F}; \ttt{Cl}; \ttt{Br}; \ttt{I}; or \ttt{None}.}
    \item{If the value of \ttt{selected} equals \ttt{true}, then the value of \ttt{element} must equal one of: \ttt{C}; \ttt{N}; \ttt{O}; \ttt{P}; or \ttt{S}.}
\end{enumerate}

Like atoms, bonds carry metadata in the form of attributes (Table \ref{table:bond-state}), which collectively define a \textit{bond state}. As with atomic attributes, not all bond states are valid. Bond state is restricted in the following ways:

\begin{table}
\caption{Bond State.}
\centering
\begin{tabular}{l L L l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{order} & Bond order & \ttt{Three} & 1 \\
    \ttt{elided} & Elision status & \ttt{boolean} & \ttt{false} \\
    \ttt{direction} & Partial conformational parity & \ttt{Option} \ttt{<Direction>} & \ttt{None} \\
    \ttt{target} & Child atom & \ttt{Atom} & - \\
    \hline
\end{tabular}
\label{table:bond-state}
\end{table}

\begin{enumerate}
    \item{If the value of \ttt{direction} does not equal \ttt{None}, then the value of \ttt{order} must equal one.}
    \item{If the value of \ttt{elided} equals \ttt{true}, the values of \ttt{order} and \ttt{direction} must equal one and \ttt{None}, respectively.}
\end{enumerate}

The data model will be referenced throughout this paper as a way to connect syntax and semantics. Dialect-compatible software may, of course, use any suitable internal data model. Implementations must, however, ensure that any internal data model is consistent with the one provided here.

\section*{Constitution}

A molecular tree is described at the lowest level by its constitution. In Dialect, constitution comprises those attributes locating nucleons and electrons at specific atoms and bonds. All other attributes lie outside the scope of constitution. Atomic attributes related to constitution are: \ttt{element}; \ttt{hydrogens}; and \ttt{charge}. Bond attributes related to constitution are \ttt{order} and \ttt{elided}.

The attributes \ttt{element} and \ttt{isotope} characterize an atom's nucleus. The \ttt{element} attribute is an optional one- or two-letter character sequence selected from the set designated by the IUPAC~\cite{periodicTable}. When IUPAC authorizes additional symbols, they will become valid values for the \ttt{element} attribute. If the elemental identity of an atom is unknown, its \ttt{element} attribute must equal \ttt{None}. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, where mass number is the sum of proton and neutron count. Setting the \ttt{isotope} property to \ttt{None} means that the element's isotopic composition equals natural abundance. Otherwise, the lower bound on the value of \ttt{isotope} is one. This lower limit allows for such physically nonsensical states such as atoms of negative implied neutron count (e.g., carbon-5). Implementations may reject such atomic states as invalid.

The \ttt{hydrogens} attribute sets an atom's mode of \textit{hydrogen suppression}. Hydrogen suppression replaces an atomic hydrogen of natural isotopic abundance and its associated bond with an integer tally associated with the parent atom. Even so, a suppressed hydrogen and its bond are understood to be present. The relative abundance and regular bonding of hydrogen atoms in molecular graphs makes hydrogen suppression an effective technique for increasing information density. Two mutually-exclusive types of suppressed hydrogen are supported: \textit{virtual hydrogen} and \textit{implicit hydrogen}.

A virtual hydrogen is one that appears as an integer unit contribution associated with the parent atom. For example, methane can be represented by a molecular graph having five nodes and four edges. But methane can also be represented as a graph of one node whose \ttt{hydrogens} attribute equals four. When using this mode, the \ttt{hydrogens} attribute may assume integer values from zero to nine, inclusive. The process of replacing atomic hydrogens with virtual hydrogen is called \enquote{virtualization.} The opposite process, replacing virtual hydrogen with atomic hydrogen, is called \enquote{reification.}

An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen node and bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as attribute on the parent atom but rather as a computation. Subject to certain limitations, an atom identifies itself as having an implicit hydrogen count by setting its \ttt{hydrogens} attribute to \ttt{Implicit}. The computation and its limitations will be described later in this paper.

The atomic attribute, \ttt{charge}, enables electron counting. The purpose of electron counting is to associate each electron in a molecular graph with either an atom or a bond. This is a fundamental requirement for the principle of mass conversion. In a more practical sense, electron counting lays the foundation for calculations of formal charge and bond order~\cite{dietz:1995}.

Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}. Although not commonly considered as such, the VB model can be thought of as an algorithm for molecular assembly. This perspective is especially useful given that Dialect itself can be viewed as a programming language for molecular assembly.

A simplified algorithm for electron counting can be summarized as follows. A molecular graph starts as a set of disconnected atoms and an integer electron count equal to the atomic number of the atom's element. Two nodes are selected as the terminals of a bond. From each atom, the same positive electron count is deducted. The total electron count deducted from each atom is simultaneously credited to the bond's electron count. For example, if each atom contributes one electron to a bond, then the bond's electron count will increase by two.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Generalized Valence Bond Electron Counting. Two unconnected nodes are chosen for bonding (left). One electron is subtracted from each atom, and both electrons are added to the bond's count (right).}
    \label{fig:electron-counting}
\end{figure}

For convenience, the actual model used by Dialect differs slightly from the above description. Neither nodes nor edges carry an explicit electron count attribute. Instead, edges carry an \ttt{order} attribute, and nodes carry a \ttt{charge} attribute. Bond order equals electron count divided by two. Formal charge equals the number of electrons gained or lost due to ionization events outside of bond formation. Gaining an electron decrements the \ttt{charge} attribute, whereas losing an electron increments it. If no electrons are gained or lost to ionization outside of bond formation, the formal charge equals zero.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Electron counting shorthand. Rather than explicit electron counts, atoms and bonds possess charge and order attributes. The atomic charge attribute captures deviations from generalized counting.}
    \label{fig:electron-counting-shorthand}
\end{figure}

From these rules follow some practical consequences. For one, bond order must be an integer greater than zero. This follows from the definition of bond formation. The electron count deducted from each atom is $n$, an integer greater than zero. Given that bond formation deducts an equal number of electrons from each atom, the total number of electrons associated with the newly-formed bond is $2n$. By definition, bond order is the bond's electron count divided by two, therefore $n$ is the bond order. Fractional, negative, and zero bond orders are all disallowed by Dialect for this reason.

Dialect imposes an upper limit of three on formal bond order. The relative scarcity of bond orders greater than this limit in molecules conforming the the VB Model makes this restriction unlikely to meaningfully restrict applicability.

Whereas negative bond order is disallowed by definition, Dialect places no restrictions on \textit{hypervalence}. Hypervalence occurs when an atom undergoes enough bonding operations to leave it with a negative implied valence electron count. Consider lithium, which possesses one valence electron. Formation of one single bond leaves lithium with zero implied valence electrons. Application of a second bond formation leaves lithium with a zero charge and an implied valence electron count of -1. Such an arrangement may be physically meaningless, but Dialect explicitly supports it. Software using Dialect may or may not reject such species for semantic reasons.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Hypervalence. Depletion of atomic electron count through excessive bonding. Hypervalent atoms are valid by default, although chemically nonsensical.}
    \label{fig:hypervalence}
\end{figure}

Each atom carries an \ttt{index} attribute, allowing bonds to reference atoms in a molecular graph. To do so, the \ttt{source} and \ttt{target} attributes are set with the respective indexes of the source and target atoms. The set of all bonds and the nodes they reference determines the connectivity of a molecular graph. In the context of determining connectivity, source and target indexes may be reversed.

\section*{Bond Elision}

A bond whose attributes are assigned default values can participate in \textit{bond elision}. In bond elision, a bond's \ttt{elided} attribute is set to \ttt{true}. When written in string form, elided bonds require no characters. This makes bond elision useful for data compression. The second application for elided bonds will be described in the next section.

\section*{Delocalization Subgraph}

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME). DIME arises when a molecular graph exists in two or more unequal forms, each differing from the others only through electron delocalization. In this context, two graphs are equal if they are isomorphic and each matched pair of atoms or bonds possess equivalent non-index attributes. If shifting electrons through delocalization causes two graphs to become equal, then DIME has occurred. DIME is sometimes referred to as \enquote{resonance} or \enquote{aromaticity,} but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.

@@

DIME can interfere with \textit{canonicalization}, or the selection of a single representation for a molecular graph. The presence of multiple equivalent molecular graphs differing only in electron delocalization complicates the formulation of selection rules and invariants, which must be adapted to account for the artificial asymmetry.

To eliminate DIME and thereby streamline canonicalization, the molecular graph is augmented with a \textit{delocalization subgraph} (DS). A DS is a possibly empty node-induced subgraph of a molecular graph. The membership of a DS may be drawn from the set of atoms and bonds that participate in DIME within a given molecular graph.

>>>>

\documentclass{article}
\usepackage{caption}
\usepackage{algorithm2e}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{array}
\usepackage{courier}
\usepackage{csquotes}
\usepackage[USenglish]{babel}
\usepackage{nameref}
\usepackage{makecell}

@@

Dialect's main goal is to solve the \textit{molecular serialization} problem. Serialization refers to the process of translating a data structure into a format that can be transmitted over a communication channel for later reconstruction. This allows general-purpose devices and networks to convey complex pieces of domain-specific information. Receiver and recipient may be separated by time, space, or both. Molecular serialization extends the concept of serialization to molecular structure. Examples of molecular serialization formats other than SMILES include Chemical Markup Language (CML)~\cite{murray-rust:2011}, CDXML~\cite{cdxml}, and CTfile (\enquote{molfile})~\cite{ctfileFormats}.

A secondary goal is brevity. Molecules should be expressable using only a few characters for each atom and either zero or one character for each bond. Brevity benefits humans who manually encode chemical structures within interactive line-driven interfaces such as those found in terminal sessions and notebooks. Brevity also benefits computers, for which compact molecular representation can improve the performance when storage, bandwidth, or processor cycles are limiting factors.

Molecular serialization schemes need to balance generality against complexity. As the diversity of the molecules that can be encoded increases, so do the rules needed to encode them and the software needed to interpret the rules. For this reason, Dialect's capabilities are directed at that region of chemical space populated by small organic molecules. Although Dialect can be used to represent a variety of other kinds of molecules, doing so in general risks data loss or corruption.

Dialect's rules can be divided into two broad categories: \textit{syntax} and \textit{semantics}. Syntax defines the set of valid strings for a language, whereas semantics determine the meaning these strings. Dialect semantic are complex due to the brevity of its syntax and the inherently complex nature of molecular representation.

Underpinning molecular serialization are conventions and abstractions for molecular representation. These are divided into three broad categories:

\begin{itemize}
\item \textit{Constitution}, the atoms present in a molecule and their bonding relationships. Dialect's extends the valence bond model with a limited form of delocalized bonding.
\item \textit{Conformation}, a kind of stereoisomerism resulting from restricted rotation about a bond. Dialect supports restricted rotation about double bonds capable of producing alkene isomerism.
\item \textit{Configuration}, a kind of stereoisomerism resulting from the arrangement of neighboring atoms in three-dimensional space. Dialect supports configurations for atoms having tetrahedral symmetry attached to four neighboring atoms.
\end{itemize}

@@

\section*{Molecular Tree}
\label{molecular-tree}

Dialect represents molecular structure as a \textit{molecular tree}. A tree is a data structure comprised of a set of nodes and pairwise relationships between them called edges. Any two nodes in a tree are connected through exactly one path of alternating nodes and edges. As such, a tree contains no cycles and exactly one connected component. A molecular tree is a specialized tree onto which chemically-relevant metadata have been overlaid. A Dialect molecule is a molecular tree having zero or more nodes and zero or more edges. The empty molecule, devoid of atoms and bonds, is therefore allowed. There is no upper bound on the number of nodes or edges, although limitations in hardware and software may restrict maximum molecular size.

A related but more common concept is the \textit{molecular graph}~\cite{balaban:1985}. Unlike molecule graphs, molecular trees do not support cycles or disconnected components. This difference might appear to make molecular trees unsuitable for molecular representation, where both rings and aggregation play important structural roles. However, these apparent shortcomings can be mitigated and even harnessed.

The main advantage of trees over graphs is information density. Graphs must accommodate cycles, which is accomplished by assigning a unique identifier to each node. An edge references two of these identifiers. Graph representations therefore incur the storage overhead of identifiers associated with both nodes and edges. In contrast, a tree can be directly encoded as a set of nested, alternating paths of nodes and edges. This simplifies the representation and eliminates the overhead of assigning and using node identifiers.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Tree vs. Graph.}
    \label{fig:tree-vs-graph}
\end{figure}

When discussing molecular trees, it's often useful to refer to \textit{node order}. An order is a property of a set that allows its members to be sorted through pairwise comparisons. Node order is the order in which the nodes of a molecular tree are added through depth-first construction. A node added before another node is said to \enquote{precede} it. A node added after another node is said to \enquote{succeed} it. In this way nodes can be prioritized. The root atom, uniquely lacking a parent, has highest priority because it precedes every other node. Similarly, a parent always precedes its children. Finally, the first child added to a given parent precedes all of its siblings. This relationship is transitive. If a node precedes (or succeeds) a sibling, then so do all of its descendants.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Node Order. Nodes in a molecular tree assume the order in which they were added through depth-first construction.}
    \label{fig:node-order}
\end{figure}

To support cycles and disconnected components, a molecular tree decouples connectivity from bonding. This means that nodes and edges don't generally map to atoms and bonds as they often do in molecular graphs. This indirection enables rings and disconnections to be encoded despite the fact that trees can contain no cycles or disconnected components.

Cycles are supported through \textit{bridges}. A bridge consists of two bond-node pairs that together represent one bond. One ach side of a bridge and atom connects to a bond, but the two atoms are not connected to each other. Instead, each atom is succeeded by a \textit{bridge node}. Unlike an atom, a bridge node has no children and so may not serve as the root of the tree. A bridge preserves the immediate atomic environment around each atom, allowing many of the same atomic computations possible in molecular graphs. Any bond, whether it closes a ring or not, can be represented as a bridge.

Each bridge node serves as a proxy to a \textit{referenced atom}. Should a bridge node be present during an operation that requires an atom, the referenced is used instead. Although a bridge node always succeeds its parent, the referenced atom may precede or succeed that parent.

Multiple molecules are supported through \textit{gaps}. A gap is an edge devoid of bonding. Gaps can be used whenever two or more molecules need to be encoded within the same molecular tree. Specific applications include ionic bonding and non-ionic complexes. However, no bonding relationship whatsoever is implied by the presence of a gap. In particular, the zero-order bond interpretation is excluded \cite{clark:2011}.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Nodes and Edge Associations. Three node/edge associations are possible within a molecular tree. A parent atom can be connected to a child atom through a bond (left). A parent atom can also be connected to a child bridge node through a bond (center). Finally, an atom can be connected, without bonding, to a child atom through a gap (right).}
    \label{fig:nodes-and-edges}
\end{figure}

Gaps and bridges decouple connectivity from bonding, but for opposite purposes. Whereas a gap connects two atoms without bonding them, a bridge creates a bond between two atoms without connecting them. These relationships can complicate the analysis of connectivity and bonding. For example, the atom succeeding a gap behaves in most situations as if it were a root node. When tallying the bonds surrounding such a node, only children are considered. Similar considerations apply when a gap succeeds an atom. Likewise, a bridge node behaves as if it were an atom in the sense that it connects to an atom through a bond. Nevertheless, a bridge node merely stands in for another atom found elsewhere in the tree. Bonding and connectivity analyses within molecular trees must therefore account for the possibility of connection-free bonds and bond-free connections.

\section*{Atoms and Bonds}
\label{atoms-and-bonds}

Within a molecular tree, atoms and bonds play complementary roles. Atoms control nuclear particle counts, valence electron counts, and configuration. Bonds determine bonding electron counts and conformation. Constitution, configuration, and conformation are all influenced by the individual and collective behavior of atoms and bonds.

At the lowest level, atom and bond behavior are governed by their respective \textit{attributes}. An attribute is a key/value pair in which the key is a name composed of a character sequence and the value is constrained by a single, immutable \textit{type}. Types can be composed and decomposed, a feature supported by many programming languages.

\begin{table}
\begin{minipage}{\textwidth}
\renewcommand*\footnoterule{}
\caption{Primitives.}
\label{table:primitives}
\centering
\begin{tabular}{l l l l}
    \hline
    Name & Description & Notation & Values \\
    \hline
    \ttt{Option} & Optional value & \ttt{\{T\}} & \ttt{None}, $t$ \\
    \ttt{List} & Ordered list of values & \ttt{[T]} & $t_0, t_1, ... t_n$ \\
    \ttt{Range} & Bounded integer range & m..n & from \ttt{m} to \ttt{n}, excluding \ttt{n} \\
    \ttt{Boolean} & Boolean value & \ttt{?} & \ttt{true}, \ttt{false} \\
    \hline
\end{tabular}
\end{minipage}
\end{table}

Dialect types are composed of four \textit{primitives} (Table~\ref{table:primitives}). The first, \ttt{Option}, represents a value of a particular type that may either be present or absent. If not present, the special value \ttt{None} is used. A \ttt{List} is a possibly empty ordered collection of values of the same type. The \ttt{Range} type represents an integer value constrained by a lower inclusive bound (\ttt{m}) and an upper, exclusive bound (\ttt{n}). The \ttt{Boolean} type assumes one of two values: \ttt{true} or \ttt{false}.

An atom is composed of seven attributes (Table~\ref{table:atomic-attributes}). Default values are applied at the time of atom creation if no other values are used. The specific values assigned to an atom collectively define an \textit{atomic state}.

\begin{table}
\caption{Atomic Attributes.}
\centering
\begin{tabular}{l L l l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{element} & Elemental symbol & \ttt{\{Element\}} & \ttt{None} \\
    \ttt{isotope} & Mass number & \ttt{\{1..1000\}} & \ttt{None} \\
    \ttt{hydrogens} & Hydrogen count & \ttt{Implicit}, \ttt{0..10} & 0 \\
    \ttt{charge} & Formal charge & \ttt{-9..10} & 0 \\
    \ttt{selected} & Selection status & \ttt{?} & \ttt{false} \\
    \ttt{parity} & Configuration & \ttt{\{AtomParity\}} & \ttt{None} \\
    \ttt{edges} & Edges directed toward child nodes & \ttt{[Edge]} & \ttt{[]} \\
    \hline
\end{tabular}
\label{table:atomic-attributes}
\end{table}

The attributes \ttt{element} (\enquote{element}) and \ttt{isotope} (\enquote{isotope}) define atomic number and neutron count, respectively. Elements are selected from a subset of those approved by IUPAC \cite{periodicTable} (Figure~\ref{fig:elements}). An atom whose element is unknown must have an \ttt{element} attribute equal to \ttt{None}. The \ttt{isotope} attribute is an optional integer value representing an atom's nuclear mass number, defined as the sum of proton and neutron count. Setting the \ttt{isotope} property to \ttt{None} means that the element's isotopic composition equals natural abundance. Otherwise, the lower bound of the value of \ttt{isotope} is one. This lower limit allows for physically nonsensical states such as atoms of negative implied neutron count (e.g., carbon-5). Implementations may reject such atomic states as invalid.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Elements. The element that may be associated with an atom.}
    \label{fig:elements}
\end{figure}

The \ttt{hydrogens} attribute sets an atom's mode of \textit{hydrogen suppression}. Hydrogen suppression replaces an atomic hydrogen of natural isotopic abundance and its associated bond with an integer tally associated with the parent atom. Even so, a suppressed hydrogen and its bond are understood to be present. The high relative abundance and regular bonding of hydrogen atoms makes hydrogen suppression an effective technique for increasing information density. Two mutually-exclusive types of suppressed hydrogen are supported: \textit{virtual hydrogen} and \textit{implicit hydrogen}.

A virtual hydrogen is one that appears as an integer unit contribution associated with the parent atom. For example, methane can be represented using five atoms and four bonds. But methane can also be represented as a single atom whose \ttt{hydrogens} attribute equals four. Virtual hydrogen count defaults to zero, and may be set to a maximum value of nine. By default, atoms use virtual hydrogen counting.

An implicit hydrogen is similar to a virtual hydrogen in that both the hydrogen node and bond have been replaced with a tally. Unlike a virtual hydrogen, however, this tally is not present as an attribute on the parent atom but rather as a computation. Setting the \ttt{hydrogens} attribute to \ttt{Implicit} activates implicit hydrogen counting. However, not all atoms are eligible to do so (see: \nameref{computing-implicit-hydrogen-count}).

The \ttt{charge} (\enquote{charge}) and \ttt{selected} (\enquote{selected}) attributes determine valence electron count (see: \nameref{electron-counting}). Charge refers formal charge, or the difference between the valence electron count of the element and the sum of bonding and nonbonding electrons of the bound atom. By default, an atom is assigned a charge of zero. Minimum and maximum values for charge are -9 and +9, respectively. The \ttt{selected} attribute adjusts an atom's selection status. Selecting an atom allows it to participate in extended bonding (see: \nameref{delocalization-subgraph}).

Configuration is determined by the \ttt{parity} attribute (\enquote{atom parity}). The type of this value, \ttt{AtomParity} is an enumeration supporting two values: \ttt{Clockwise} and \ttt{Counterclockwise} (Table~\ref{table:atom-parity-variants}). The interpretation and encoding of atom parity will be described in detail later (see: \nameref{configuration}).

\begin{table}
\caption{Atom Parity Variants.}
\centering
\begin{tabular}{l l}
    \hline
    Variant & Description  \\
    \hline
    \ttt{Clockwise} & Configuration runs with node order. \\
    \ttt{Counterclockwise} & Configuration runs against node order. \\
    \hline
\end{tabular}
\label{table:atom-parity-variants}
\end{table}

The \ttt{edges} ("edges") attribute is a possibly empty ordered list of edges associated with an atom. As noted in the previous section, an edge may represent either a bond or a gap (Table~\ref{table:edge-variants}). The mere presence of an edge does not guarantee a bonding relationship.

\begin{table}
\caption{Edge Variants.}
\centering
\begin{tabular}{l l}
    \hline
    Variant & Description  \\
    \hline
    \ttt{Bond} & A bonding and connectivity relationship. \\
    \ttt{Gap} & A connectivity relationship only. \\
    \hline
\end{tabular}
\label{table:edge-variants}
\end{table}

Not all atomic states are valid. Implementations must ensure either the impossibility of creating an invalid state, or an error condition in the event that one is created. Atomic state is restricted in the following two ways:

\begin{enumerate}
    \item{If implicit hydrogen counting is used, default values of isotope, charge, and parity must also be used. Implicit hydrogen counting is only supported for elements in the list: B; C, N; O; P; S; F; Cl; Br; or I.}
    \item{A selected atom must use one of the following five elements: C; N; O; P; or S.}
\end{enumerate}

A bond is composed of four attributes (Table~\ref{table:bond-attributes}). The \ttt{order} attribute (\enquote{order}) represents the concept of formal bond order, or the number of bonding electrons divided by two. A bond's order may only assume the values one, two, or three. The \ttt{elided} attribute (\enquote{elided}) allows a bond to be omitted during serialization and participate in extended bonding. The \ttt{direction} attribute (\enquote{direction}) is used with other information to conformationally restrict a bond. The two allowed variants for \ttt{Direction}, \ttt{Up} and \ttt{Down}, will be described in detail later (see: \nameref{conformation}). The \ttt{target} attribute identifies the child node at which the bond is directed (Table~\ref{table:target-variants}).

\begin{table}
\caption{Bond Attributes.}
\centering
\begin{tabular}{l p{3.6cm} l l}
    \hline
    Attribute & Description & Type & Default \\
    \hline
    \ttt{order} & Formal bond order & \ttt{1..4} & 1 \\
    \ttt{elided} & Elision status & \ttt{boolean} & \ttt{false} \\
    \ttt{direction} & Partial conformational parity & \ttt{\{Direction\}} & \ttt{None} \\
    \ttt{target} & An atom or bridge node & \ttt{Target} & - \\
    \hline
\end{tabular}
\label{table:bond-attributes}
\end{table}

\begin{table}
\caption{Target Variants.}
\centering
\begin{tabular}{l p{6.6cm}}
    \hline
    Variant & Description \\
    \hline
    \ttt{Atom} & A parent and child atom are bound. \\
    \tt{Bridge} & A parent atom and bridge node are bound. \\
    \hline
\end{tabular}
\label{table:target-variants}
\end{table}

As with atomic state, not all bond states are valid. The following restrictions apply:

\begin{enumerate}
    \item{An elided bond must have an order of one and default direction.}
    \item{A bond with a non-default direction must have an order of one.}
\end{enumerate}

The atom and bond attributes defined here will be referenced throughout this paper as a way to connect syntax and semantics. Implementations may, of course, use any suitable internal data model. They must, however, ensure that any internal data model is consistent with the one provided here.

\section*{Electron Counting}
\label{electron-counting}

A molecular tree explicitly encodes the presence of protons and neutrons through the atomic attributes \ttt{element} and \ttt{isotope}, respectively. The presence of electrons, in contrast, is merely implied. The principles of charge and mass conservation nevertheless require a method to determine explicit electron counts. This is possible through \textit{electron counting}.

Electron counting in Dialect is based on the well-known \textit{valence bond model} (VB Model)~\cite{lewis:1916}. This model assumes that a bond spans exactly two atoms and is associated with a nonzero, positive, and even number of electrons. These electrons are drawn in equal quantities from each atom's set of valence electrons. A single bond draws one electron from each atom (two total), a double bond two (four total), and a triple bond three (six total). The electron count of an atom whose \ttt{element} attribute equals \ttt{None} is always zero. Bonding to such an atom adds the required number of electrons directly to the bond.

Electrons are counted as follows. A molecular tree starts as a single root atom. The electron count of this atom equals the difference between proton count (atomic number) and charge. Adding a bond is a three-step process. First, a child atom is constructed with an electron count determined in the same way as the root atom. Next, the child atom and the bond are added to the molecular tree. Finally, the electron counts of each atom and the bond are updated. If parent and child are connected through a gap edge, no electrons are deducted from either atom. Should a parent atom be bound to a child bridge node, then the electron count of the referenced atom is reduced.

\begin{figure}
    \centering
    \includegraphics{filler}
    \caption{Electron Counting. A root atom and its child are added to a tree (left). Next, a bond is added (center). Finally, the appropriate electron count is deducted from each atom and added to the bond (right).}
    \label{fig:electron-counting}
\end{figure}

Whereas negative bond order is disallowed by definition, Dialect places no restrictions on \textit{hypervalence}. Hypervalence occurs when an atom undergoes enough bonding operations to leave it with a negative implied valence electron count. Consider lithium, which possesses one valence electron. Formation of one single bond leaves lithium with zero implied valence electrons. Application of a second bond formation leaves lithium with a zero charge and an implied valence electron count of -1. Such an arrangement may be physically meaningless, but Dialect explicitly supports it. Software using Dialect may or may not reject such species on semantic grounds.

\section*{Delocalization Subgraph}
\label{delocalization-subgraph}

A molecular representation based solely on the valence bond model can yield artifacts resulting from \textit{delocalization induced molecular equality} (DIME). DIME arises when a given molecule can be represented by two forms, each differing from the other only in double bond placement. DIME is sometimes referred to as \enquote{resonance} or \enquote{aromaticity,} but those terms are avoided here due to their ambiguity~\cite{randic:2018,kerber:2006}.

@@

The main problem with DIME is its interference with \textit{canonicalization}, or the selection of a single molecular representation among many alternatives. The presence of multiple equivalent molecular forms differing only in electron delocalization complicates the formulation of selection rules and invariants, which must be adapted to account for the artificial asymmetry.

To eliminate DIME and thereby streamline canonicalization, each molecular tree is augmented with a \textit{delocalization subgraph} (DS). A DS is a possibly empty node-induced subgraph over a molecular tree. The membership of a DS will typically be drawn from the set of atoms and bonds that participate in DIME within a given molecular tree.

Membership of a DS is determined as follows. An atom is added to the DS by setting its \ttt{selected} attribute to \ttt{true} through a process called \textit{selection}. As noted previously (see: \nameref{atoms-and-bonds}), only some atomic states are compatible with selection. An edge will be added to the DS if both of the atoms it spans are selected and the bond itself is elided. This allows arbitrary bonds between selected atoms to be excluded from the DS. If a bond's target is a bridge node, the corresponding edge will be added to the DS if both the parent atom and the equivalent target atom are selected.
